Guarded Recursion in Agda via Sized Types

### 2019-04-18 code-oneclock-paper/latex/Introduction.tex
Small fixes

Possibly non-terminating computations
and infinite structures, such as non-wellfounded trees, can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
need to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, recursive
calls must appear directly under
the application of a constructor. A similar syntactic
check is used in Coq. % \cite{DBertotK08}.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques to ensure the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
 is a type annotated with the number of unfolding that
elements of this type can undergo \cite{Hughes96}. Sized types have been implemented
in Agda and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns,AVW-normalization}, as exemplified
by Abel and Chapman's work \cite{AC14}.

Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types.
These types can be used in combination with clock quantification to define  coinductive types \cite{atkey2013productive,bizjak2016guarded,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.
\remove{, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.}
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal\ \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

Afterwards we develop a categorical semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically the semantics
of type theories for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic,MannaaM18,Mogelberg14}.
Here, to clarify the connection between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types. 
The well-order relation on sizes is fundamental for constructing a
terminating definition of the semantic fixpoint combinator.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-gtt}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

>>>>

Possibly non-terminating computations
and infinite structures, such as non-wellfounded trees, can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
need to be productive in the sense that every
finite part of the output can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, recursive
calls must appear directly under
the application of a constructor. A similar syntactic
check is used in Coq. % \cite{DBertotK08}.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques to ensure the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
 is a type annotated with the number of unfolding that
elements of this type can undergo \cite{Hughes96}. Sized types have been implemented
in Agda and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns,AVW-normalization}, as exemplified
by Abel and Chapman's work \cite{AC14}.
There also are other approaches to sized types, one of those is developed by Sacchini
\cite{Sacchini13}, but they allow less terms to be typed compared to Agda.

Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types.
These types can be used in combination with clock quantification to define  coinductive types \cite{atkey2013productive,BahrGM17,bizjak2016guarded}. Currently there is no
implementation of a calculus for guarded recursion.
\remove{, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.}
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aforementioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \ \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. 
%In this respect \GTT\ is closer in spirit to Atkey and McBride's
%language.
The design of \GTT\ has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

Afterwards we develop a categorical semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically, the semantics
of a type theory for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic,MannaaM18,Mogelberg14}.
Here, to clarify the connection between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types. 
The well-order relation on sizes is fundamental for constructing a
terminating definition of the semantic fixpoint combinator.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-gtt}. The formalization
uses Agda 2.5.4.2 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code passed Agda's type and termination checker.

### 2019-04-18 code-oneclock-paper/latex/paper.tex
Small fixes

\begin{abstract}
  In type theory, programming and reasoning with possibly
  non-terminating programs and potentially infinite objects is
  achieved using coinductive types. Recursively defined programs of
  these types need to be productive to guarantee the
  consistency of the type system. Proof assistants such as Agda and
  Coq traditionally employ strict syntactic productivity checks, which
  commonly make programming with coinductive types convoluted.  One
  way to overcome this issue is by encoding productivity at the level
  of types so that the type system forbids the implementation of
  non-productive corecursive programs.
%%, meaning
%%  that although their definitions can be unfolded \emph{ad infinitum},
%%  finite unfoldings are guaranteed to produce . 
  In this paper we compare two different approaches to type-based
  productivity: guarded recursion and sized types. More specifically,
  we show how to simulate guarded recursion in Agda using sized
  types. We formalize the syntax of a simple type theory for guarded
  recursion, which is a variant of Atkey and McBride's
  calculus for productive coprogramming. Then we give a denotational semantics
  using presheaves over the preorder of sizes. Sized types are
  fundamentally used to interpret the characteristic
  features of guarded recursion, notably the fixpoint combinator.
%Sized types are fundamentally employed in the interpretation of the characteristic features of guarded recursion, notably the fixpoint combinator.
%%extending type theory with
%%  coinduction and coinductive types.  More specifically, we show that
%%  guarded recursion and guarded recursive types can be simulated with
%%  sized types.  We do this by defining a model of guarded recursion in
%%  type theory extended with sized types.  The model is variation of
%%  the topos of trees, and it is formalized in Agda.
\end{abstract}

@@

We also need to provide a semantic version of 
%To interpret simple type theory, we also need to give 
the context operations, the simple type formers and the operations on substitutions.
Since their definitions are standard, we do not discuss them in detail.
For each of them, we need to make a case distinction based on the
state of the clock context.
For example, the empty variable context \IC{•} is interpreted
as the unit type in the clock context \IC{∅}, and it is interpreted as 
the terminal presheaf in the clock context \IC{κ}.
%, depending on the state of the clock context.
We use the operations on presheaves defined in \Cref{sec:presheaves} to interpret simple type formers.
\input{CloTT/TypeFormers/FunctionType}

>>>>

\begin{abstract}
  In type theory, programming and reasoning with possibly
  non-terminating programs and potentially infinite objects is
  achieved using coinductive types. Recursively defined programs of
  these types need to be productive to guarantee the
  consistency of the type system. Proof assistants such as Agda and
  Coq traditionally employ strict syntactic productivity checks, which
  often make programming with coinductive types convoluted.  One
  way to overcome this issue is by encoding productivity at the level
  of types so that the type system forbids the implementation of
  non-productive corecursive programs.
%%, meaning
%%  that although their definitions can be unfolded \emph{ad infinitum},
%%  finite unfoldings are guaranteed to produce . 
  In this paper we compare two different approaches to type-based
  productivity: guarded recursion and sized types. More specifically,
  we show how to simulate guarded recursion in Agda using sized
  types. We formalize the syntax of a simple type theory for guarded
  recursion, which is a variant of Atkey and McBride's
  calculus for productive coprogramming. Then we give a denotational semantics
  using presheaves over the preorder of sizes. Sized types are
  fundamentally used to interpret the characteristic
  features of guarded recursion, notably the fixpoint combinator.
%Sized types are fundamentally employed in the interpretation of the characteristic features of guarded recursion, notably the fixpoint combinator.
%%extending type theory with
%%  coinduction and coinductive types.  More specifically, we show that
%%  guarded recursion and guarded recursive types can be simulated with
%%  sized types.  We do this by defining a model of guarded recursion in
%%  type theory extended with sized types.  The model is variation of
%%  the topos of trees, and it is formalized in Agda.
\end{abstract}

@@

We also need to provide a semantic version of 
%To interpret simple type theory, we also need to give 
the context operations, the simple type formers and the operations on substitutions.
Since their definitions are standard, we do not discuss them in detail.
For each of them, we need to make a case distinction based on the clock context.
For example, the empty variable context \IC{•} is interpreted
as the unit type in the clock context \IC{∅}, and it is interpreted as 
the terminal presheaf in the clock context \IC{κ}.
%, depending on the state of the clock context.
We use the operations on presheaves defined in \Cref{sec:presheaves} to interpret simple type formers.
\input{CloTT/TypeFormers/FunctionType}
